The Genetic Algorithm of GMSE

### 2021-07-30 notebook/cran_vignettes/SI1.Rmd
Update GMSE version and equation, checking off an item in Issue #64 for the documentation

---
title: 'The Genetic Algorithm of GMSE'
subtitle: 'GMSE: an R package for generalised management strategy evaluation (Supporting Information 1)'
author: |
        | A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;, Jeroen Minderman&#xb9;, 
        | Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy Rakotonarivo&#xb9;, 
        | Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
date: '[2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
output:
  html_document: default
  pdf_document:
    citation_package: natbib
    fig_caption: yes
  word_document:
    fig_caption: yes
    reference_docx: docx_template.docx
header-includes:
- \usepackage{lineno}
- \linenumbers
linestretch: 1
bibliography: gmse.bib
biblio-style: apalike
link-citations: true
linkcolor: blue
---

@@

A genetic algorithm is called in the predefined GMSE manager and user models to simulate human decision making. As of GMSE version 0.6, this includes one independent call to the genetic algorithm for each decision-making agent in every GMSE time step. Therefore, one run of the genetic algorithm occurs to simulate the manager's policy-setting decisions in each time step (unless otherwise defined through non-default `manage_freq` values greater than 1; e.g., see [Management frequency and extinction risk](SI6.html)), and one run occurs to simulate each individual user's action decisions in each time step (unless otherwise defined through non-default `group_think = TRUE`, in which case one user makes decisions that all other users copy). Each run of the genetic algorithm mimics the evolution by natural selection of a population of potential manager or user strategies over multiple iterations, with the highest fitness strategy in the terminal iteration being selected as the one that the manager or user decides to implement. For clarity, as in the main text, we use 'time step' to refer to a full GMSE cycle (in which multiple genetic algorithms may be run) and 'iteration' to refer to a single, non-overlapping, generation of potential strategies that evolve within a genetic algorithm (see Figure 1 of the [main text](ms.pdf)). Below, we explain the genetic algorithm in detail, as it occurs in GMSE v0.6 (future versions of GMSE might expand upon this framework, and we highlight some of these potential avenues for expansion). We first explain the [key data structures used](#Key_data_structures_used), then provide an overview of how a population of strategies is [initialised](#initialised), and the subsequent processes of [crossover](#crossover), [mutation](#mutation), [cost constraint](#constraint), [fitness](#eval) evaluation, [tournament selection](#tournament), and [replacement](#replacement). We then explain the fitness functions of [managers](#manfit) and [users](#ufit) in more detail.

@@

The focal data structure used for tracking manager and user decisions is a three dimensional array, which we will call `ACTION` (also returned as `user_array` by `gmse_apply`; see [Default GMSE data structure](SI7.html)). Rows of `ACTION` correspond to the entities affected by actions (resources, landscape properties, or potentially other agents), and columns correspond either to properties of the affected entities, or to the actions potentially allocated to them. Each layer of `ACTION` corresponds to a unique agent, the first of which is the manager; additional layers correspond to users. Below shows an `ACTION` array for a GMSE model with one manager and two users.

@@

Columns 2-4 refer to resource or landscape types, but only `Type_1 = 1`, `Type_2 = 0`, and `Type_3 = 0 ` are allowed in predefined GMSE v0.6 manager and user sub-models (i.e., only one type of resource is permitted). Future versions might allow for different resource types (e.g., `Type_1` might be used to designate species, and `Type_2` and `Type_3` could designate stage or sex). Column 5 `Util.` of `ACTION` defines the utility associated with the resource (where `Act = -2`) or landscape (where `Act = -1`). For managers, the target resource abundance set with the GMSE argument `manage_target` is found in row 1 (`r samp_ACTION[1,5,1]` in `ACTION` above); for users, the value in row 1 identifies whether resources are preferred to increase (if positive) or decrease (if negative). Values of column 5 in row 2 similarly identify whether landscape cell output is preferred by users to increase or decrease (managers do not currently have preferences for landscape output). Of special note is row 3 for `Manager_Actions`, which defines the *current* manager's utility for resources; that is, the adjustment to resource abundance that the manager will attempt to make based on the `manage_target` and the most recent estimate of resource abundance produced by the observation model (in the case of the above, resource abundance is estimated at ca `r round((samp_ACTION[1,5,1] + -1*samp_ACTION[3,5,1]),2)`, so the manager will set policy in attempt to change the population size by ca `r round(samp_ACTION[3,5,1],2)` resources). Column 6 `U_land` defines whether or not the utility attached to the resource or landscape output depends on it being on a landscape cell that is owned by the acting user. Related, column 7 `U_loc.` defines whether or not actions can be performed only on a landscape cell that is owned by the acting user. Hence values of columns 6 and 7 are binary, and affected by the `land_ownership` argument in `gmse` and `gmse_apply`. Finally, columns 8-13 correspond to specific actions, either direct (where `Act < 0`) or indirect by setting policy (for row 3 of `Manager_Actions` where `Act = 1`). The last column 13 `None` corresponds with no actions. See [GMSE documentation](https://cran.r-project.org/web/packages/GMSE/GMSE.pdf) for details about the effects of each action.

Constraints on the values that elements in the `ACTION` array can take are defined by a `COST` array (also returned as `manager_array` by `gmse_apply`; see [Default GMSE data structures](SI7.html)) of dimensions identical to `ACTION`. Elements of `COST` define how many units from the `manager_budget` or `user_budget` are needed to perform a single action; a `minimum_cost` for actions is defined as an argument in GMSE (10 by default). All values in `COST` columns 1-7 are set to 100001, one higher than the highest possible `manager_budget` or `user_budget`, so neither managers nor users can affect resource types or utilities. Columns 8-13 are also set to 100001, except where actions are allowed. Maximum values of 100000 are independent of any other parameter value specified in GMSE (e.g., landscape dimensions). Below shows the `COST` array that corresponds to the above `ACTION` array.

@@

Here we explain how the fitnesses of candidate manager and user strategies in a `POPULATION` array (see above) are calculated. We emphasise that the fitness functions used in GMSE v0.6 are intended to be heuristic tools for identifying reasonable manager and user behaviours. In practice, our fitness functions identify behaviours that are well-aligned with manager and user interests for harvesting or crop yield, but they are not intended to identify *optimal* decisions. This practical, metaheuristic approach is consistent with the objectives of management strategy evaluation [@Bunnefeld2011], and is well-suited for the use of genetic algorithms [@Hamblin2013]. @Luke2009 describes the metaheuristic approach more generally (original emphasis retained):

@@

Individual fitness as calculated for managers ($F_{i}^{m}$) is affected by a manager's utility for resources and the projected change in resource abundance caused by the individual's policy (i.e., the contents of their `POPULATION` layer, specifically row 3; here again we use 'individual' to refer to one of `ga_popsize` discrete strategies in `POPULATION`, which may be selected and reproduce within the genetic algorithm). Manager utility for a resource ($U^{m}_{res}$) is defined as the difference between `manage_target` and the estimation of population abundance as produced by the GMSE observation model (see "[Key data structures used](#Key_data_structures_used)" above, and [Default GMSE data structures](SI7.html#action) for more information). Manager utility can therefore change in each GMSE time step as estimated resource abundance changes; when the estimated resource abundance is greater than `manage_target`, $U^{m}_{res}$ is negative, and when the estimated resource abundance is less than `manage_target`, $U^{m}_{res}$ is positive. To get the fitness of individuals, first the change in resource abundance predicted by the individual's policy ($\Delta A_{i}$) is calculated, then the squared difference between $\Delta A_{i}$ and $U^{m}_{res}$ is calculated to obtain a utility deviation ($D_{i}$) for the individual $i$,

@@

The predicted consequences of $X_{i,j}$ for resource abundance differ for each possible action. For each action, no consequence is predicted if the policy is not allowed by a simulation of GMSE (e.g., `culling = FALSE`). For allowed actions, the parameter `manager_sense` ($\sigma$) modulates predicted consequences for abundance by some factor; this is useful because not all actions attempted by users will be realised, and a value of $\sigma = 1$ tends to slightly overestimate how much the actions attempted by users will actually translate to a change in resource abundance. In practice, the default $\sigma = 0.9$ usually performs well. Managers also have access to an estimate of the number of offspring that resources produce ($E[offspring]$). For default settings in GMSE v0.6, $E[offspring] = \lambda$, where $\lambda$ is the GMSE argument `lambda` that defines the baseline population growth rate of resources. Given a non-default GMSE argument for `consume_repr` (where `consume_repr` is how much consumption of landscape yield is needed to increment offspring production by 1), $E[offspring]$ is further incremented by `( (times_feeding * res_consume) / consume_repr)` to account for reproduction enabled by feeding on the landscape (`res_consume` is the proportion of landscape yield that a resource consumes during feeding).

@@

Note that $F_{i}^{u}$ increases when $\Delta A_{i}$ and $\Delta L_{i}$ are of the same sign as $U^{u}_{res}$ and $U^{u}_{land}$, respectively. Further, in GMSE v0.6, only one term of the equation is nonzero. When `land_ownership = FALSE` (default, modelling users that harvest resources), $U^{u}_{res} = -1$ and $U^{u}_{land} = 0$, and when `land_ownership = TRUE`, $U^{u}_{res} = 0$ and $U^{u}_{land} = 100$ (modelling farmers trying to increase crop yield). Hence users only have a single objective of either decreasing resource abundance or increasing landscape output, though landscape output might be increased indirectly by decreasing resource abundance if `res_consume > 0`.

@@

$$\Delta A_{i} = (\lambda) Feeds + Helps -Scares - Culls - (\lambda) Castrations.$$

>>>>

---
title: 'The Genetic Algorithm of GMSE'
subtitle: 'GMSE: an R package for generalised management strategy evaluation (Supporting Information 1)'
author: |
        | A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;, Jeroen Minderman&#xb9;, 
        | Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy Rakotonarivo&#xb9;, 
        | Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling, UK [2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
  html_document: default
  word_document:
    fig_caption: yes
    reference_docx: docx_template.docx
header-includes:
- \usepackage{lineno}
linestretch: 1
bibliography: gmse.bib
biblio-style: apalike
link-citations: true
linkcolor: blue
---

@@

A genetic algorithm is called in the predefined GMSE manager and user models to simulate human decision making. As of GMSE version 0.6, this includes one independent call to the genetic algorithm for each decision-making agent in every GMSE time step. Therefore, one run of the genetic algorithm occurs to simulate the manager's policy-setting decisions in each time step (unless otherwise defined through non-default `manage_freq` values greater than 1; e.g., see [Management frequency and extinction risk](SI6.pdf)), and one run occurs to simulate each individual user's action decisions in each time step (unless otherwise defined through non-default `group_think = TRUE`, in which case one user makes decisions that all other users copy). Each run of the genetic algorithm mimics the evolution by natural selection of a population of potential manager or user strategies over multiple iterations, with the highest fitness strategy in the terminal iteration being selected as the one that the manager or user decides to implement. For clarity, as in the main text, we use 'time step' to refer to a full GMSE cycle (in which multiple genetic algorithms may be run) and 'iteration' to refer to a single, non-overlapping, generation of potential strategies that evolve within a genetic algorithm (see Figure 1 of the [main text](ms.pdf)). Below, we explain the genetic algorithm in detail, as it occurs in GMSE v0.6+ (future versions of GMSE might expand upon this framework, and we highlight some of these potential avenues for expansion). We first explain the [key data structures used](#Key_data_structures_used), then provide an overview of how a population of strategies is [initialised](#initialised), and the subsequent processes of [crossover](#crossover), [mutation](#mutation), [cost constraint](#constraint), [fitness](#eval) evaluation, [tournament selection](#tournament), and [replacement](#replacement). We then explain the fitness functions of [managers](#manfit) and [users](#ufit) in more detail.

@@

The focal data structure used for tracking manager and user decisions is a three dimensional array, which we will call `ACTION` (also returned as `user_array` by `gmse_apply`; see [Default GMSE data structure](SI7.pdf)). Rows of `ACTION` correspond to the entities affected by actions (resources, landscape properties, or potentially other agents), and columns correspond either to properties of the affected entities, or to the actions potentially allocated to them. Each layer of `ACTION` corresponds to a unique agent, the first of which is the manager; additional layers correspond to users. Below shows an `ACTION` array for a GMSE model with one manager and two users.

@@

Columns 2-4 refer to resource or landscape types, but only `Type_1 = 1`, `Type_2 = 0`, and `Type_3 = 0 ` are allowed in predefined GMSE v0.6+ manager and user sub-models (i.e., only one type of resource is permitted). Future versions might allow for different resource types (e.g., `Type_1` might be used to designate species, and `Type_2` and `Type_3` could designate stage or sex). Column 5 `Util.` of `ACTION` defines the utility associated with the resource (where `Act = -2`) or landscape (where `Act = -1`). For managers, the target resource abundance set with the GMSE argument `manage_target` is found in row 1 (`r samp_ACTION[1,5,1]` in `ACTION` above); for users, the value in row 1 identifies whether resources are preferred to increase (if positive) or decrease (if negative). Values of column 5 in row 2 similarly identify whether landscape cell output is preferred by users to increase or decrease (managers do not currently have preferences for landscape output). Of special note is row 3 for `Manager_Actions`, which defines the *current* manager's utility for resources; that is, the adjustment to resource abundance that the manager will attempt to make based on the `manage_target` and the most recent estimate of resource abundance produced by the observation model (in the case of the above, resource abundance is estimated at ca `r round((samp_ACTION[1,5,1] + -1*samp_ACTION[3,5,1]),2)`, so the manager will set policy in attempt to change the population size by ca `r round(samp_ACTION[3,5,1],2)` resources). Column 6 `U_land` defines whether or not the utility attached to the resource or landscape output depends on it being on a landscape cell that is owned by the acting user. Related, column 7 `U_loc.` defines whether or not actions can be performed only on a landscape cell that is owned by the acting user. Hence values of columns 6 and 7 are binary, and affected by the `land_ownership` argument in `gmse` and `gmse_apply`. Finally, columns 8-13 correspond to specific actions, either direct (where `Act < 0`) or indirect by setting policy (for row 3 of `Manager_Actions` where `Act = 1`). The last column 13 `None` corresponds with no actions. See [GMSE documentation](https://cran.r-project.org/web/packages/GMSE/GMSE.pdf) for details about the effects of each action.

Constraints on the values that elements in the `ACTION` array can take are defined by a `COST` array (also returned as `manager_array` by `gmse_apply`; see [Default GMSE data structures](SI7.pdf)) of dimensions identical to `ACTION`. Elements of `COST` define how many units from the `manager_budget` or `user_budget` are needed to perform a single action; a `minimum_cost` for actions is defined as an argument in GMSE (10 by default). All values in `COST` columns 1-7 are set to 100001, one higher than the highest possible `manager_budget` or `user_budget`, so neither managers nor users can affect resource types or utilities. Columns 8-13 are also set to 100001, except where actions are allowed. Maximum values of 100000 are independent of any other parameter value specified in GMSE (e.g., landscape dimensions). Below shows the `COST` array that corresponds to the above `ACTION` array.

@@

Here we explain how the fitnesses of candidate manager and user strategies in a `POPULATION` array (see above) are calculated. We emphasise that the fitness functions used in GMSE v0.6+ are intended to be heuristic tools for identifying reasonable manager and user behaviours. In practice, our fitness functions identify behaviours that are well-aligned with manager and user interests for harvesting or crop yield, but they are not intended to identify *optimal* decisions. This practical, metaheuristic approach is consistent with the objectives of management strategy evaluation [@Bunnefeld2011], and is well-suited for the use of genetic algorithms [@Hamblin2013]. @Luke2009 describes the metaheuristic approach more generally (original emphasis retained):

@@

Individual fitness as calculated for managers ($F_{i}^{m}$) is affected by a manager's utility for resources and the projected change in resource abundance caused by the individual's policy (i.e., the contents of their `POPULATION` layer, specifically row 3; here again we use 'individual' to refer to one of `ga_popsize` discrete strategies in `POPULATION`, which may be selected and reproduce within the genetic algorithm). Manager utility for a resource ($U^{m}_{res}$) is defined as the difference between `manage_target` and the estimation of population abundance as produced by the GMSE observation model (see "[Key data structures used](#Key_data_structures_used)" above, and [Default GMSE data structures](SI7.pdf) for more information). Manager utility can therefore change in each GMSE time step as estimated resource abundance changes; when the estimated resource abundance is greater than `manage_target`, $U^{m}_{res}$ is negative, and when the estimated resource abundance is less than `manage_target`, $U^{m}_{res}$ is positive. To get the fitness of individuals, first the change in resource abundance predicted by the individual's policy ($\Delta A_{i}$) is calculated, then the squared difference between $\Delta A_{i}$ and $U^{m}_{res}$ is calculated to obtain a utility deviation ($D_{i}$) for the individual $i$,

@@

The predicted consequences of $X_{i,j}$ for resource abundance differ for each possible action. For each action, no consequence is predicted if the policy is not allowed by a simulation of GMSE (e.g., `culling = FALSE`). For allowed actions, the parameter `manager_sense` ($\sigma$) modulates predicted consequences for abundance by some factor; this is useful because not all actions attempted by users will be realised, and a value of $\sigma = 1$ tends to slightly overestimate how much the actions attempted by users will actually translate to a change in resource abundance. In practice, the default $\sigma = 0.9$ usually performs well. Managers also have access to an estimate of the number of offspring that resources produce ($E[offspring]$). For default settings in GMSE v0.6+, $E[offspring] = \lambda$, where $\lambda$ is the GMSE argument `lambda` that defines the baseline population growth rate of resources. Given a non-default GMSE argument for `consume_repr` (where `consume_repr` is how much consumption of landscape yield is needed to increment offspring production by 1), $E[offspring]$ is further incremented by `( (times_feeding * res_consume) / consume_repr)` to account for reproduction enabled by feeding on the landscape (`res_consume` is the proportion of landscape yield that a resource consumes during feeding).

@@

Note that $F_{i}^{u}$ increases when $\Delta A_{i}$ and $\Delta L_{i}$ are of the same sign as $U^{u}_{res}$ and $U^{u}_{land}$, respectively. Further, in GMSE v0.6+, only one term of the equation is nonzero. When `land_ownership = FALSE` (default, modelling users that harvest resources), $U^{u}_{res} = -1$ and $U^{u}_{land} = 0$, and when `land_ownership = TRUE`, $U^{u}_{res} = 0$ and $U^{u}_{land} = 100$ (modelling farmers trying to increase crop yield). Hence users only have a single objective of either decreasing resource abundance or increasing landscape output, though landscape output might be increased indirectly by decreasing resource abundance if `res_consume > 0`.

@@

$$\Delta A_{i} = (\lambda) Feeds + Helps - Scares - (1 + \lambda)Culls - (\lambda) Castrations.$$

### 2021-07-30 notebook/cran_vignettes/SI7.Rmd
Update GMSE version and equation, checking off an item in Issue #64 for the documentation

Many data structures contain elements that are unused in GMSE v0.6, and in all cases this is designed for ease of ongoing development of new GMSE features. Requests for new features can be made on GitHub using the [GMSE Wiki](https://github.com/bradduthie/gmse/wiki) or the [GMSE Issues](https://github.com/bradduthie/gmse/issues) page.

>>>>

Many data structures contain elements that are unused in GMSE v0.6+, and in all cases this is designed for ease of ongoing development of new GMSE features. Requests for new features can be made on GitHub using the [GMSE Wiki](https://github.com/bradduthie/gmse/wiki) or the [GMSE Issues](https://github.com/bradduthie/gmse/issues) page.

### 2021-07-30 notebook/website_vignettes/SI1.Rmd
Update GMSE version and equation, checking off an item in Issue #64 for the documentation

A genetic algorithm is called in the predefined GMSE manager and user models to simulate human decision making. As of GMSE version 0.6, this includes one independent call to the genetic algorithm for each decision-making agent in every GMSE time step. Therefore, one run of the genetic algorithm occurs to simulate the manager's policy-setting decisions in each time step (unless otherwise defined through non-default `manage_freq` values greater than 1; e.g., see [Management frequency and extinction risk](SI6.html)), and one run occurs to simulate each individual user's action decisions in each time step (unless otherwise defined through non-default `group_think = TRUE`, in which case one user makes decisions that all other users copy). Each run of the genetic algorithm mimics the evolution by natural selection of a population of potential manager or user strategies over multiple iterations, with the highest fitness strategy in the terminal iteration being selected as the one that the manager or user decides to implement. For clarity, as in the main text, we use 'time step' to refer to a full GMSE cycle (in which multiple genetic algorithms may be run) and 'iteration' to refer to a single, non-overlapping, generation of potential strategies that evolve within a genetic algorithm (see Figure 1 of the [main text](ms.pdf)). Below, we explain the genetic algorithm in detail, as it occurs in GMSE v0.6 (future versions of GMSE might expand upon this framework, and we highlight some of these potential avenues for expansion). We first explain the [key data structures used](#Key_data_structures_used), then provide an overview of how a population of strategies is [initialised](#initialised), and the subsequent processes of [crossover](#crossover), [mutation](#mutation), [cost constraint](#constraint), [fitness](#eval) evaluation, [tournament selection](#tournament), and [replacement](#replacement). We then explain the fitness functions of [managers](#manfit) and [users](#ufit) in more detail.

@@

Columns 2-4 refer to resource or landscape types, but only `Type_1 = 1`, `Type_2 = 0`, and `Type_3 = 0 ` are allowed in predefined GMSE v0.6 manager and user sub-models (i.e., only one type of resource is permitted). Future versions might allow for different resource types (e.g., `Type_1` might be used to designate species, and `Type_2` and `Type_3` could designate stage or sex). Column 5 `Util.` of `ACTION` defines the utility associated with the resource (where `Act = -2`) or landscape (where `Act = -1`). For managers, the target resource abundance set with the GMSE argument `manage_target` is found in row 1 (`r samp_ACTION[1,5,1]` in `ACTION` above); for users, the value in row 1 identifies whether resources are preferred to increase (if positive) or decrease (if negative). Values of column 5 in row 2 similarly identify whether landscape cell output is preferred by users to increase or decrease (managers do not currently have preferences for landscape output). Of special note is row 3 for `Manager_Actions`, which defines the *current* manager's utility for resources; that is, the adjustment to resource abundance that the manager will attempt to make based on the `manage_target` and the most recent estimate of resource abundance produced by the observation model (in the case of the above, resource abundance is estimated at ca `r round((samp_ACTION[1,5,1] + -1*samp_ACTION[3,5,1]),2)`, so the manager will set policy in attempt to change the population size by ca `r round(samp_ACTION[3,5,1],2)` resources). Column 6 `U_land` defines whether or not the utility attached to the resource or landscape output depends on it being on a landscape cell that is owned by the acting user. Related, column 7 `U_loc.` defines whether or not actions can be performed only on a landscape cell that is owned by the acting user. Hence values of columns 6 and 7 are binary, and affected by the `land_ownership` argument in `gmse` and `gmse_apply`. Finally, columns 8-13 correspond to specific actions, either direct (where `Act < 0`) or indirect by setting policy (for row 3 of `Manager_Actions` where `Act = 1`). The last column 13 `None` corresponds with no actions. See [GMSE documentation](https://cran.r-project.org/web/packages/GMSE/GMSE.pdf) for details about the effects of each action.

>>>>

A genetic algorithm is called in the predefined GMSE manager and user models to simulate human decision making. As of GMSE version 0.6, this includes one independent call to the genetic algorithm for each decision-making agent in every GMSE time step. Therefore, one run of the genetic algorithm occurs to simulate the manager's policy-setting decisions in each time step (unless otherwise defined through non-default `manage_freq` values greater than 1; e.g., see [Management frequency and extinction risk](SI6.html)), and one run occurs to simulate each individual user's action decisions in each time step (unless otherwise defined through non-default `group_think = TRUE`, in which case one user makes decisions that all other users copy). Each run of the genetic algorithm mimics the evolution by natural selection of a population of potential manager or user strategies over multiple iterations, with the highest fitness strategy in the terminal iteration being selected as the one that the manager or user decides to implement. For clarity, as in the main text, we use 'time step' to refer to a full GMSE cycle (in which multiple genetic algorithms may be run) and 'iteration' to refer to a single, non-overlapping, generation of potential strategies that evolve within a genetic algorithm (see Figure 1 of the [main text](ms.pdf)). Below, we explain the genetic algorithm in detail, as it occurs in GMSE v0.6+ (future versions of GMSE might expand upon this framework, and we highlight some of these potential avenues for expansion). We first explain the [key data structures used](#Key_data_structures_used), then provide an overview of how a population of strategies is [initialised](#initialised), and the subsequent processes of [crossover](#crossover), [mutation](#mutation), [cost constraint](#constraint), [fitness](#eval) evaluation, [tournament selection](#tournament), and [replacement](#replacement). We then explain the fitness functions of [managers](#manfit) and [users](#ufit) in more detail.

@@

Columns 2-4 refer to resource or landscape types, but only `Type_1 = 1`, `Type_2 = 0`, and `Type_3 = 0 ` are allowed in predefined GMSE v0.6+ manager and user sub-models (i.e., only one type of resource is permitted). Future versions might allow for different resource types (e.g., `Type_1` might be used to designate species, and `Type_2` and `Type_3` could designate stage or sex). Column 5 `Util.` of `ACTION` defines the utility associated with the resource (where `Act = -2`) or landscape (where `Act = -1`). For managers, the target resource abundance set with the GMSE argument `manage_target` is found in row 1 (`r samp_ACTION[1,5,1]` in `ACTION` above); for users, the value in row 1 identifies whether resources are preferred to increase (if positive) or decrease (if negative). Values of column 5 in row 2 similarly identify whether landscape cell output is preferred by users to increase or decrease (managers do not currently have preferences for landscape output). Of special note is row 3 for `Manager_Actions`, which defines the *current* manager's utility for resources; that is, the adjustment to resource abundance that the manager will attempt to make based on the `manage_target` and the most recent estimate of resource abundance produced by the observation model (in the case of the above, resource abundance is estimated at ca `r round((samp_ACTION[1,5,1] + -1*samp_ACTION[3,5,1]),2)`, so the manager will set policy in attempt to change the population size by ca `r round(samp_ACTION[3,5,1],2)` resources). Column 6 `U_land` defines whether or not the utility attached to the resource or landscape output depends on it being on a landscape cell that is owned by the acting user. Related, column 7 `U_loc.` defines whether or not actions can be performed only on a landscape cell that is owned by the acting user. Hence values of columns 6 and 7 are binary, and affected by the `land_ownership` argument in `gmse` and `gmse_apply`. Finally, columns 8-13 correspond to specific actions, either direct (where `Act < 0`) or indirect by setting policy (for row 3 of `Manager_Actions` where `Act = 1`). The last column 13 `None` corresponds with no actions. See [GMSE documentation](https://cran.r-project.org/web/packages/GMSE/GMSE.pdf) for details about the effects of each action.
